Modelling fine-scale dive behavior with hidden Markov models

### 2020-07-06 Report/Sections/results.tex

To apply this methodology to real-world data, the CarHHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada. The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 hertz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. A plot of the processed data for all dives and one selected dive can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.

@@

Information criteria tends to overestimate the number of states in biological processes \cite{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily. This is a common inssue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria. Section \ref{subsec:model_validation} describes our process of validating this model in particular.

@@

\subsection{Model evaluation}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
	\caption{Psuedoresiduals of $Y^{*(2)}$}
	\label{fig:pseudoresids}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
	\caption{Empirical Distribution of $Y^*(1)_x$ for sub-dive state 3 plotted against its estimated pdf}
	\label{fig:empirical_dist}
\end{figure}

>>>>

To apply this methodology to real-world data, the CarHHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada. The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 hertz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. A plot of the processed data for all dives and a collection of five selected dive can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.

\begin{figure}[h!]
	\centering
	\includegraphics[height=5in]{../Plots/lagplot.png}
	\caption{Lag plot of vertical velocity and $\tilde a$ (left) and a the associated normal kernel density estimates (right)}
	\label{fig:data}
\end{figure}

\begin{figure}[h!]
	\centering
	\includegraphics[height=5in]{../Plots/lagplot.png}
	\caption{Lag plot of vertical velocity and $\tilde a$ (left) and a the associated normal kernel density estimates (right)}
	\label{fig:data_one_dive}
\end{figure}

@@

Information criteria tends to overestimate the number of states in biological processes \cite{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily. This is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria. Section \ref{subsec:model_validation} describes our process of validating this model in particular.

@@

\subsection{Model Validation}
\label{model_validation}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=5.5in]{../Plots/pseudoresids.png}
	\caption{Psuedoresiduals of $Y^{*(2)}$}
	\label{fig:pseudoresids}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=5.5in]{../Plots/empirical_dist.png}
	\caption{Empirical Distribution of $Y^{*(1)}_z$ for sub-dive state 3 plotted against its estimated pdf}
	\label{fig:empirical_dist}
\end{figure}
